---
title: "knitr_project"
author: "iskren89"
date: "7/5/2020"
output:
  html_document: default
  pdf_document: default
---
# Reproducible research project
  
**1. Here we are loading and preprocessing the data**  
```{r}
library(dplyr)
library(knitr)
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
my_file <- unz(temp, "activity.csv")
dat <- read.csv(my_file)
dat[,3]<-factor(dat[,3])
missing<-sum(is.na(dat$steps))
unlink(temp)
```
  
**2. Here we plot the total number of steps taken each day**   
```{r fig.width=13}
by_date<-group_by(dat,date)
total_date<-summarise(by_date,sum=sum(steps),.groups = 'keep')
with(total_date, hist(sum,main="Total number of steps",xlab=""))
```
  
**3. Here we calculate the mean and median number of steps taken each day**  
```{r}
summary(total_date)
```
  
**4. Here we show the time series plot of the average number of steps taken**   
```{r fig.width=13}
by_interval<-group_by(dat,interval)
mean_interval<-summarise(by_interval,mean=mean(steps,na.rm=TRUE),.groups = 'keep')
with(mean_interval, plot(interval,mean,type = "l",main="Average number of steps",ylab="steps by 5-minute interval"))
```
  
**5. The 5-minute interval that, on average, contains the maximum number of step is:**  
```{r}
mean_interval[which.max(mean_interval$mean),]
```
  
**6. Here we describe and show a strategy for imputing missing data**  
There are `r missing` missing values in the original dataset - there are 8 days for which there is no data.  
We will impute the missing data by using the mean for that 5-minute interval.  
```{r}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
imputed<-mutate(by_interval,steps = impute.mean(steps))
```
  
**7. Here we show a histogram of the total number of steps taken each day after missing values are imputed**   
```{r fig.width=13}
imputed_by_date<-group_by(imputed,date)
imputed_total_date<-summarise(imputed_by_date,sum=sum(steps),.groups = 'keep')
with(imputed_total_date, hist(sum,main="Total number of steps",xlab=""))
```
  
We see that the profile of the data does not change after we impute it.  
```{r}
summary(imputed_total_date)
```
  
*Note that in the original calculation and histogram I do not remove the 8 days for which we have no data.*  
*If you use na.rm=TRUE to remove the missing values, they are still counted as 0 in any summary statistics.*  
*As a result the mean/median total number of steps per day will be lower, because the 8 missing days are all taken as 0 in the numerator, while the denominator is higher.*  
*I took this approach because I think it's unrealistic that no steps were taken on those 8 days.*  
*Rather the more plausible explanation is that the person did not wear his or hers activity tracker on that day. The effect of this can be seen below.*   
  
*Missing data not removed from the original data set. Note that there are 8 missing values and the minimum number of steps is 41:*  
```{r} 
summary(total_date)
```
  
*Missing data removed from the original data set. Note that there are no missing values and the minimum number of steps is 0:*  
```{r}
removed_total_date<-summarise(by_date,sum=sum(steps,na.rm=TRUE),.groups = 'keep')
summary(removed_total_date)
```
  
**8. Here we make a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**  
```{r}
imputed$date<-as.Date(imputed$date)
imputed$weekday<-as.factor(weekdays(imputed$date))
weekend <- c("Saturday","Sunday")
imputed$weekend <- as.factor(ifelse(imputed$weekday %in% weekend, "weekend","weekday"))
imputed_weekdays<-filter(imputed,weekend=="weekday")
imputed_weekends<-filter(imputed,weekend=="weekend")
weekdays_interval<-group_by(imputed_weekdays,interval)
interval_weekdays<-summarise(weekdays_interval,mean=mean(steps,na.rm=TRUE),.groups = 'keep')
weekends_interval<-group_by(imputed_weekends,interval)
interval_weekends<-summarise(weekends_interval,mean=mean(steps,na.rm=TRUE),.groups = 'keep')
par(mfrow=c(2,1), mar=c(4,2,1,1))
with(interval_weekdays, plot(interval,mean,type = "l",main="Average number of steps in weekdays"))
with(interval_weekends, plot(interval,mean,type = "l",main="Average number of steps in weekends"))
```